Descriptive Complexity for Counting Complexity Classes

### 2018-12-09 LMCS/beyond.tex
all direct replacements by reviewer 1

%!TEX root = main.tex

We have used weighted logics to give a framework for descriptive complexity of counting complexity classes. Here, we go beyond weighted logics and give the first steps on defining recursion at the quantitative level.
This goal is not trivial not only because we want to add recursion over functions, but also because 
it is not clear what could be the right notion of ``fixed point''. 
To this end, we show first how to extend $\qso$ with function symbols that are later
used 
to define a natural generalization for functions of 
the notion of least fixed point of LFP.
 As a proof of concept, we show that this notion can be used to captures $\fp$.
Moreover, we use this concept
to define an operator for counting paths in a graph, a natural generalization of the transitive closure operator~\cite{immerman1999descriptive}, and show that this gives rise to a logic that captures~$\shl$.

@@

As for the case of LFP, it would be natural to consider the point-wise partial order $\leq$ on $\F$ defined as $f \leq g$ if, and only if, $f(i) \leq g(i)$ for every $i \in \{1, \ldots, \ell\}$, and let the semantics of $\clfp{\beta(\x,h)}$ be the least fixed point of the operator $T_\beta$. However, $(\F, \leq)$ is not a complete lattice, so we do not have a Knaster-Tarski Theorem ensuring that such a fixed point exists. Instead, we generalize the semantics of LFP as follows. In the definition of the semantics of LFP, an operator $T$ on relations is considered, and the semantics is defined in terms of the least fixed point of $T$, that is, a relation $R$ such that~\cite{I86,vardi1982complexity}: 
(a) $T(R) = R$, and (b) $R \subseteq S$ for every $S$ such that $T(S) = S$. 
We can view $T$ as an operator on functions if we consider the characteristic function of a relation. Given a relation $R \subseteq A^\ell$, let $\chi_R$ be its characteristic function, that is $\chi_R(\bar a) = 1$ if $\bar a \in R$, and $\chi_R(\bar a) = 0$ otherwise. Then define an operator $T^\star$ on characteristic functions as $T^\star(\chi_R) = \chi_{T(R)}$. Moreover, we can rewrite the conditions defining a least fixed point of $T$ in terms of the operator $T^\star$ if we consider the notion of support of a function. Given a function $f \in \F$, define the support of $f$, denoted by $\support(f)$, as $\{ \bar a \in A^\ell \mid f(\bar a) > 0 \}$. Then given that $\support(\chi_R) = R$, we have that the conditions (a) and (b) are equivalent to the following conditions on $T^\star$:
(a) $\support(T^\star(\chi_R)) = \support(\chi_R)$, and  (b) $\support(\chi_R) \subseteq \support(\chi_S)$ for every $S$ such that  $\support(T^\star(\chi_{S})) = \support(\chi_S)$.
To define a notion of fixed point for $T_\beta$ we simply generalized these conditions. More precisely, a function $f \in \F$ is a {\em s-fixed point} of $T_{\beta}$ if $\support(T_\beta(f)) = \support(f)$, and $f$ is a {\em least s-fixed point} of $T_{\beta}$ if $f$ is a s-fixed point of $T_\beta$ and for every s-fixed point $g$ of $T_\beta$ it holds that $\support(f) \subseteq \support(g)$. The existence of such fixed point is ensured by the following lemma:
\begin{lem}\label{lem-support}
Let $h \in \fs$ such that $\arity(h) = \ell$, and $\beta$ be an $\fqfo(\fo)$-formula over a signature $\R$ such that if a function symbol occurs in $\beta$, then this function symbol is $h$. Moreover, let $\A$ be an $\R$-structure with domain $A$, $f,g : A^\ell \to \mathbb{N}$ and $F,G$ be function assignments such that $F(h) = f$ and $F(h) = g$. If $\support(f) \subseteq \support(g)$, then for every first-order and second-order assignments $v$ and $V$, respectively, it holds that:
\begin{center}
if $\sem{\beta}(\A,v,V,F) > 0$, then $\sem{\beta}(\A,v,V,G) > 0$.
\end{center}
\end{lem}
\proof
\input{proof-lem-support}
\qed

In the particular case of an $\rqfo(\fo)$-formula $\clfp{\beta(\x, h)}$, Lemma  \ref{lem-support} tell us that if $f,g \in \F$ and $\support(f) \subseteq \support(g)$, then $\support(T_\beta(f)) \subseteq \support(T_\beta(g))$. Hence, as for the case of LFP, this lemma gives us a simple way to compute a least s-fixed point of $T_\beta$. Let $f_0 \in \F$ be a function such that $f_0(\bar a) = 0$ for every $\bar a \in A^\ell$ (i.e. $f_0$ is the only function with empty support), and let function $f_{i+1}$ be defined as $T_\beta(f_i)$ for every $i \in \N$. Then there exists $j \geq 0$ such that $\support(f_j) = \support(T_\beta(f_j))$. Let $k$ be the smallest natural number such that $\support(f_{k}) = \support(T_\beta(f_k))$. We have that $f_k$ is a least s-fixed point of $T_\beta$, which is used to defined the semantics of $\clfp{\beta(\x, h)}$. More specifically, for an arbitrary first-order assignment $v$ for $\A$:
\begin{align*}
\sem{\clfp{\beta(\x, h)}}(\A,v) &= f_{k}(v(\x)).
\end{align*}

@@

Assume now that we need to extend our previous counting function to the case of arbitrary directed graphs. To this end, suppose that $\varphi_{\text{\rm first}}(x)$ and $\varphi_{\text{succ}}(x,y)$ are $\fo$-formulae defining the first element of $<$ and the successor relation associated to $<$, respectively.\footnote{Recall that in this paper we consider ordered $\R$-structures.} Moreover, define formula $\beta(x, y, t, g)$ as follows:
\begin{align*}
(E(x,y) + \sa{z} g(x,z,t)\cdot E(z,y)) \cdot \varphi_{\text{\rm first}}(t) \ +
\sa{t'} \varphi_{\text{succ}}(t',t) \cdot \left(\sa{x'} \sa{y'} g(x',y',t') \right)
\end{align*}
Then our extended counting function is defined by:
$$
\sa{t} (\varphi_{\text{\rm first}}(t) \cdot \clfp{\beta(x,y,t,g)}).
$$ 
In fact, the number of paths of length at most $n$ from a node $x$ to a node $y$ is recursively computed by using the formula $(E(x,y) + \sa{z} g(x,z,t)\cdot E(z,y)) \cdot \varphi_{\text{\rm first}}(t)$, which stores this value in $g(x,y,t)$ with $t$ the first element in the domain.  The other formula $\sa{t'} \varphi_{\text{succ}}(t',t) \cdot \left(\sa{x'} \sa{y'} g(x',y',t') \right)$ is just an auxiliary artifact that is used as a counter to allow reaching a fixed point in the support of $g$ in $n$ steps. Notice that the use of the filter $\varphi_{\text{succ}}(t',t)$ prevents this formula for incrementing the value of $g(x,y,t)$ when $t$ is the first element in the domain. \qed
\end{exa}
In contrast with $\lfp$, to reach a fixed point we do not need to impose any positive restriction on the formula $\beta(\x,h)$.
In fact, since $\beta$ is constructed from monotones operations (sum and product) over the natural numbers, the resulting operator $T_{\beta}$ is monotone as well.

@@

Given a relation signature $\R$, the set of transitive $\qfo$ formulae ($\tqfo$-formulae) is defined as an extension of $\qfo$ with the formula $[\pth \psi(\bar{x},\bar{y})]$, where $\psi(\x, \y)$ is an $\fo$-formula over $\R$, and $\bar{x} = (x_1, \ldots, x_k)$, $\bar{y} = (y_1, \ldots, y_k)$ are tuples of pairwise distinct first-order variables. The semantics of $[\pth \psi(\bar{x},\bar{y})]$ can easily be defined in terms of $\rqfo(\fo)$ as follows. 

@@

As it was mentioned before, the operator for counting paths is exactly what we need to capture $\shl$.

@@

As a example showing how fundamental this separation is, consider the issue of extending $\qfo(\fo)$ at the Boolean level in order to capture $\shl$. The natural alternative to do this is to use $\fo$ extended with a transitive closure operator, which is denoted by $\tc$. But then the problem is that for every language $L \in \nlog$, it holds that its characteristic function $\chi_L$ is in $\qfo(\tc)$, where $\chi_L(x) = 1$ if $x \in L$, and $\chi_L(x) = 0$ otherwise. Thus, if we assume that $\qfo(\tc)$ captures $\shl$ (over ordered structures), then we have that $\chi_L \in \shl$ for every $L \in \nlog$. This would imply that $\nlog = \ulog$,\footnote{A decision language $L$ is in $\ulog$ is there exists a logarithmic-space NTM $M$ accepting $L$ and satisfying that $\tma_M(x) = 1$ for every $x \in L$.} solving an outstanding open problem \cite{Reinhardt97}.

>>>>

% !TeX root = main.tex

We have used weighted logics to give a framework for descriptive complexity of counting complexity classes. Here, we go beyond weighted logics and give the first steps on defining recursion at the quantitative level.
This goal is not trivial not only because we want to add recursion over functions, but also because 
it is not clear what could be the right notion of ``fixed point''. 
To this end, we show first how to extend $\qso$ with function symbols that are later
used 
to define a natural generalization for functions of 
the notion of least fixed point of LFP.
 As a proof of concept, we show that this notion can be used to capture $\fp$.
Moreover, we use this concept
to define an operator for counting paths in a graph, a natural generalization of the transitive closure operator~\cite{immerman1999descriptive}, and show that this gives rise to a logic that captures~$\shl$.

@@

As for the case of LFP, it would be natural to consider the point-wise partial order $\leq$ on $\F$ defined as $f \leq g$ if, and only if, $f(\bar{a}) \leq g(\bar{a})$ for every $\bar{a} \in A^{\ell}$, and let the semantics of $\clfp{\beta(\x,h)}$ be the least fixed point of the operator $T_\beta$. However, $(\F, \leq)$ is not a complete lattice, so we do not have a Knaster-Tarski Theorem ensuring that such a fixed point exists. Instead, we generalize the semantics of LFP as follows. In the definition of the semantics of LFP, an operator $T$ on relations is considered, and the semantics is defined in terms of the least fixed point of $T$, that is, a relation $R$ such that~\cite{I86,vardi1982complexity}: 
(a) $T(R) = R$, and (b) $R \subseteq S$ for every $S$ such that $T(S) = S$. 
We can view $T$ as an operator on functions if we consider the characteristic function of a relation. Given a relation $R \subseteq A^\ell$, let $\chi_R$ be its characteristic function, that is $\chi_R(\bar a) = 1$ if $\bar a \in R$, and $\chi_R(\bar a) = 0$ otherwise. Then define an operator $T^\star$ on characteristic functions as $T^\star(\chi_R) = \chi_{T(R)}$. Moreover, we can rewrite the conditions defining a least fixed point of $T$ in terms of the operator $T^\star$ if we consider the notion of support of a function. Given a function $f \in \F$, define the support of $f$, denoted by $\support(f)$, as $\{ \bar a \in A^\ell \mid f(\bar a) > 0 \}$. Then given that $\support(\chi_R) = R$, we have that the conditions (a) and (b) are equivalent to the following conditions on $T^\star$:
(a) $\support(T^\star(\chi_R)) = \support(\chi_R)$, and  (b) $\support(\chi_R) \subseteq \support(\chi_S)$ for every $S$ such that  $\support(T^\star(\chi_{S})) = \support(\chi_S)$.
To define a notion of fixed point for $T_\beta$ we simply generalize these conditions. More precisely, a function $f \in \F$ is a {\em s-fixed point} of $T_{\beta}$ if $\support(T_\beta(f)) = \support(f)$, and $f$ is a {\em least s-fixed point} of $T_{\beta}$ if $f$ is a s-fixed point of $T_\beta$ and for every s-fixed point $g$ of $T_\beta$ it holds that $\support(f) \subseteq \support(g)$. The existence of such fixed point is ensured by the following lemma:
\begin{lem}\label{lem-support}
Let $h \in \fs$ such that $\arity(h) = \ell$, and $\beta$ be an $\fqfo(\fo)$-formula over a signature $\R$ such that if a function symbol occurs in $\beta$, then this function symbol is $h$. Moreover, let $\A$ be an $\R$-structure with domain $A$, $f,g : A^\ell \to \mathbb{N}$ and $F,G$ be function assignments such that $F(h) = f$ and $F(h) = g$. If $\support(f) \subseteq \support(g)$, then for every first-order and second-order assignments $v$ and $V$, respectively, it holds that:
\begin{center}
if $\sem{\beta}(\A,v,V,F) > 0$, then $\sem{\beta}(\A,v,V,G) > 0$.
\end{center}
\end{lem}
\proof
\input{proof-lem-support}
\qed

In the particular case of an $\rqfo(\fo)$-formula $\clfp{\beta(\x, h)}$, Lemma  \ref{lem-support} tell us that if $f,g \in \F$ and $\support(f) \subseteq \support(g)$, then $\support(T_\beta(f)) \subseteq \support(T_\beta(g))$. Hence, as for the case of LFP, this lemma gives us a simple way to compute a least s-fixed point of $T_\beta$. Let $f_0 \in \F$ be a function such that $f_0(\bar a) = 0$ for every $\bar a \in A^\ell$ (i.e. $f_0$ is the only function with empty support), and let function $f_{i+1}$ be defined as $T_\beta(f_i)$ for every $i \in \N$. Then there exists $j \geq 0$ such that $\support(f_j) = \support(T_\beta(f_j))$. Let $k$ be the smallest natural number such that $\support(f_{k}) = \support(T_\beta(f_k))$. We have that $f_k$ is a least s-fixed point of $T_\beta$, which is used to define the semantics of $\clfp{\beta(\x, h)}$. More specifically, for an arbitrary first-order assignment $v$ for $\A$:
\begin{align*}
\sem{\clfp{\beta(\x, h)}}(\A,v) &= f_{k}(v(\x)).
\end{align*}

@@

Assume now that we need to extend our previous counting function to the case of arbitrary directed graphs. To this end, suppose that $\varphi_{\text{\rm first}}(x)$ and $\varphi_{\text{succ}}(x,y)$ are $\fo$-formulae defining the first element of $<$ and the successor relation associated to $<$, respectively.\footnote{Recall that in this paper we consider ordered $\R$-structures.} Moreover, define formula $\beta(x, y, t, g)$ as follows:
\begin{align*}
(E(x,y) + \sa{z} g(x,z,t)\cdot E(z,y)) \cdot \varphi_{\text{\rm first}}(t) \ +
\sa{t'} \varphi_{\text{succ}}(t',t) \cdot \left(\sa{x'} \sa{y'} g(x',y',t') \right)
\end{align*}
Then our extended counting function is defined by:
$$
\sa{t} (\varphi_{\text{\rm first}}(t) \cdot \clfp{\beta(x,y,t,g)}).
$$ 
In fact, the number of paths of length at most $n$ from a node $x$ to a node $y$ is recursively computed by using the formula $(E(x,y) + \sa{z} g(x,z,t)\cdot E(z,y)) \cdot \varphi_{\text{\rm first}}(t)$, which stores this value in $g(x,y,t)$ with $t$ the first element in the domain.  The other formula $\sa{t'} \varphi_{\text{succ}}(t',t) \cdot \left(\sa{x'} \sa{y'} g(x',y',t') \right)$ is just an auxiliary artifact that is used as a counter to allow reaching a fixed point in the support of $g$ in $n$ steps. Notice that the use of the filter $\varphi_{\text{succ}}(t',t)$ prevents this formula for incrementing the value of $g(x,y,t)$ when $t$ is the first element in the domain. \qed
\end{exa}
In contrast to $\lfp$, to reach a fixed point we do not need to impose any positive restriction on the formula $\beta(\x,h)$.
In fact, since $\beta$ is constructed from monotone operations (sum and product) over the natural numbers, the resulting operator $T_{\beta}$ is monotone as well.

@@

Given a relational signature $\R$, the set of transitive $\qfo$ formulae ($\tqfo$-formulae) is defined as an extension of $\qfo$ with the formula $[\pth \psi(\bar{x},\bar{y})]$, where $\psi(\x, \y)$ is an $\fo$-formula over $\R$, and $\bar{x} = (x_1, \ldots, x_k)$, $\bar{y} = (y_1, \ldots, y_k)$ are tuples of pairwise distinct first-order variables. The semantics of $[\pth \psi(\bar{x},\bar{y})]$ can easily be defined in terms of $\rqfo(\fo)$ as follows. 

@@

As mentioned before, the operator for counting paths is exactly what we need to capture $\shl$.

@@

As an example showing how fundamental this separation is, consider the issue of extending $\qfo(\fo)$ at the Boolean level in order to capture $\shl$. The natural alternative to do this is to use $\fo$ extended with a transitive closure operator, which is denoted by $\tc$. But then the problem is that for every language $L \in \nlog$, it holds that its characteristic function $\chi_L$ is in $\qfo(\tc)$, where $\chi_L(x) = 1$ if $x \in L$, and $\chi_L(x) = 0$ otherwise. Thus, if we assume that $\qfo(\tc)$ captures $\shl$ (over ordered structures), then we have that $\chi_L \in \shl$ for every $L \in \nlog$. This would imply that $\nlog = \ulog$,\footnote{A decision language $L$ is in $\ulog$ is there exists a logarithmic-space NTM $M$ accepting $L$ and satisfying that $\tma_M(x) = 1$ for every $x \in L$.} solving an outstanding open problem \cite{Reinhardt97}.

### 2018-12-09 LMCS/extentions.tex
all direct replacements by reviewer 1

%!TEX root = main.tex

There exist complexity classes that do not fit in our framework because either the output of a function in not a natural number (e.g. a negative number) or the class is not defined purely in terms of arithmetical operations (e.g. min and max).
To remedy this problem, we show here how $\qso$ can be easily extended  to capture such classes that go beyond sum and product over natural numbers.

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{FFK94} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{exa}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. This can be easily done in $\qsoz$ with the formula:
	$
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1 
	$.
\end{exa}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{cor} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{cor}
This is an interesting result that shows how robust and versatile is $\qso$ for capturing different counting complexity classes even beyond $\bbN$.

A different class of functions comes from considering the optimization version of a decision problem. For example, one can define MAX-SAT as the problem of determining the maximum number of clauses, of a given CNF propositional formula, that can be made true by an assignment. Here, MAX-SAT is defined in terms of a maximization problem which in its essence differs from the functions in $\shp$. 
To formalize this set of optimization problems, Krentel defined $\optp$~\cite{krentel1988complexity} as the class of functions computable by taking the maximum or minimum of the output values over all runs of a polynomial-time NTM machine with output tape (i.e. each run produces a binary string which is interpreted as a number). 
For instance, MAX-SAT is in $\optp$ as many other optimization versions of $\np$-problems.
Given that in~\cite{krentel1988complexity} Krentel did not make the distinction between $\max$ and $\min$, in~\cite{vollmer1995complexity} they defined the classes $\maxp$ and $\minp$ as the max and min version of the problems in $\optp$ (i.e. $\optp = \maxp \cup \minp$).

In order to capture classes of optimization functions, we extend as follows $\qso$ with $\max$ and $\min$ quantifiers (called $\optqso$). 
Given a signature $\R$, the set of $\optqso$-formulae over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{align*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{align*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing, respectively, instead of computing a sum or a product. 
\begin{exa}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the largest clique in a graph. This can be done in $\optqso$ as follows:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
	Notice that formula $\sa{z} X(z)$ is used to compute the number of nodes in a set $X$. 
\end{exa}
Similar than for $\maxp$ and $\minp$, we have to distinguished between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulae constructed from $\qfo$ operators and $\max$-formulae $\max\{\alpha,\alpha\}$, $\maxa{x} \alpha$ and  $\maxa{X} \alpha$.
The class $\minqso$ is defined analogously changing $\max$ with $\min$. Notice that in $\maxqso$ and $\minqso$, second-order sum and product are not allowed. For instance, formula $\alpha_6$ in Example \ref{ex:optqso} is in $\maxqso$.
As one could expect, $\maxqso$ and $\minqso$ are the needed logics to capture $\maxp$ and $\minp$.
\begin{thm} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ capture $\maxp$ and $\minp$, respectively, over ordered structures.
\end{thm}
\proof
\input{proof-capture-optp}
\qed
It is important to mention that a similar result, following the framework of~\cite{SalujaST95}, was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ (resp., $\minpb$) of problems in $\maxp$ (resp., $\minp$) whose output value is polynomially bounded.
Interestingly, Theorem \ref{theo:capture-optp} is stronger since our logic has the freedom to use sum and product quantifiers, instead of using a max-and-count problem over Boolean formulae. 
Finally, it is easy to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.

>>>>

% !TeX root = main.tex

There exist complexity classes that do not fit in our framework because either the output of a function is not a natural number (e.g. a negative number) or the class is not defined purely in terms of arithmetical operations (e.g. min and max).
To remedy this problem, we show here how $\qso$ can be easily extended  to capture such classes that go beyond sum and product over natural numbers.

It is well-known that, under some reasonable complexity-theoretical assumptions, $\shp$ is not closed under subtraction, not even under subtraction by one~\cite{OH93}.
To overcome this limitation, $\gp$ was introduced in~\cite{FFK94} as the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x) - \tmr_M(x)$, where  $\tmr_M(x)$ is the number of rejecting runs of $M$ with input $x$.
That is, $\gp$ is the closure of $\shp$ functions under subtraction, and its functions can obviously take negative values.
Given that our logical framework was built on top of the natural numbers, we need to extend $\qso$ in order to capture $\gp$. 
The most elegant way to do this is by allowing constants coming from $\bbZ$ instead of just $\bbN$. 
Formally, we define the logic $\qsoz$ whose syntax is the same as in \eqref{syntax} and whose semantics is the same as in Table~\ref{tab-semantics} except that the atomic formula $s$ (i.e. a constant) comes from $\bbZ$.  
Similar than for $\qso$, we define the fragment $\eqsoz$ as the extension of $\eqso$ with constants in $\bbZ$.
\begin{exa}
	Recall the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the number of cliques in a graph that are not triangles. This can be easily done in $\qsoz$ with the formula:
	$
	\alpha_5 :=	\alpha_2 + (-1) \cdot \alpha_1. 
	$ \qed
\end{exa}
Adding negative constants is a mild extension to allow subtraction in the logic. 
It follows from our characterization of $\shp$ that this is exactly what we need to capture  $\gp$.
\begin{cor} \label{prop:capture-gapp}
	$\eqsoz(\fo)$ captures $\gp$ over ordered structures.
\end{cor}
This is an interesting result that shows how robust and versatile $\qso$ is for capturing different counting complexity classes even beyond $\bbN$.

A different class of functions comes from considering the optimization version of a decision problem. For example, one can define MAX-SAT as the problem of determining the maximum number of clauses, of a given CNF propositional formula that can be made true by an assignment. Here, MAX-SAT is defined in terms of a maximization problem which in its essence differs from the functions in $\shp$. 
To formalize this class of optimization problems, Krentel defined $\optp$~\cite{krentel1988complexity} as the class of functions computable by taking the maximum or minimum of the output values over all runs of a polynomial-time NTM machine with output tape (i.e. each run produces a binary string which is interpreted as a number). 
For instance, MAX-SAT is in $\optp$ as many other optimization versions of $\np$-problems.
Given that in~\cite{krentel1988complexity} Krentel did not make the distinction between $\max$ and $\min$, in~\cite{vollmer1995complexity} they defined the classes $\maxp$ and $\minp$ as the max and min version of the problems in $\optp$ (i.e. $\optp = \maxp \cup \minp$).

In order to capture classes of optimization functions, we extend $\qso$ with $\max$ and $\min$ quantifiers as follows (called $\optqso$). 
Given a signature $\R$, the set of $\optqso$-formulae over $\R$ is given by extending the syntax in (\ref{syntax}) with the following operators:
\begin{align*}
\max\{\alpha,\alpha\} \ \mid\ \min\{\alpha,\alpha\} \ \mid \maxa{x} \alpha \ \mid \ \mina{x} \alpha \ \mid \ \maxa{X} \alpha \ \mid \ \mina{X} \alpha 
\end{align*}
where $x \in \fv$ and $X \in \sv$. The semantics of the $\qso$-operators in $\optqso$ are defined as usual. Furthermore, the semantics of the max and min quantifiers are defined as the natural extension of the sum quantifiers in $\qso$ (see Table~\ref{tab-semantics}) by maximizing or minimizing, respectively, instead of computing a sum or a product. 
\begin{exa}\label{ex:optqso}
	Recall again the setting of Example~\ref{ex:cliques} and suppose now that we want to compute the size of the largest clique in a graph. This can be done in $\optqso$ as follows:
	\[
\alpha_6 := \maxa{X} \left( \, \clique(X) \cdot \sa{z} X(z)  \, \right)
	\]
	Notice that formula $\sa{z} X(z)$ is used to compute the number of nodes in a set $X$.  \qed
\end{exa}
Similar than for $\maxp$ and $\minp$, we have to distinguish between the $\max$ and $\min$ fragments of $\optqso$. For this, we define the fragment $\maxqso$ of all $\optqso$ formulae constructed from $\qfo$ operators and $\max$-formulae $\max\{\alpha,\alpha\}$, $\maxa{x} \alpha$ and  $\maxa{X} \alpha$.
The class $\minqso$ is defined analogously replacing $\max$ by $\min$. Notice that in $\maxqso$ and $\minqso$, second-order sum and product are not allowed. For instance, formula $\alpha_6$ in Example \ref{ex:optqso} is in $\maxqso$.
As one could expect, $\maxqso$ and $\minqso$ are the needed logics to capture $\maxp$ and $\minp$.
\begin{thm} \label{theo:capture-optp}
	$\maxqso(\fo)$ and $\minqso(\fo)$ capture $\maxp$ and $\minp$, respectively, over ordered structures.
\end{thm}
\proof
\input{proof-capture-optp}
\qed
It is important to mention that a similar result, following the framework of~\cite{SalujaST95}, was proved in~\cite{kolaitis1994logical} for the class $\maxpb$ (resp., $\minpb$) of problems in $\maxp$ (resp., $\minp$) whose output value is polynomially bounded.
Interestingly, Theorem \ref{theo:capture-optp} is stronger since our logic has the freedom to use sum and product quantifiers, instead of using a max-and-count problem over Boolean formulae. 
Finally, it is easy to prove that our framework can also capture $\maxpb$ and $\minpb$ by disallowing the product $\Pi x$ in $\maxqso(\fo)$ and $\minqso(\fo)$, respectively.
